Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
try to refactor section 3 to contain less repetition and to fix the "backjumping" that toon disliked (and that michael complained about too).

The \emph{code generation} phase takes as input the trace generated during
\emph{tracing}.

@@

When a hot loop is identified, the interpreter enters a
special mode, called \emph{tracing mode}. During tracing, the interpreter
records a history of all the operations it executes.

Such a history is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by generating
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code. To ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example conditions and indirect or virtual
calls.  When generating the machine code, every guard is turned into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit the machine code and continue the execution by falling
back to interpretation.

\toon{that part seems a bit out of place, it jumps back to a point in the story
before what you explained right before it. (as far as I can tell)}
During tracing, the trace is repeatedly
checked as to whether the interpreter is at a position in the program where it had been
earlier. If this happens, the trace recorded corresponds to a loop
in the interpreted program. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be executed immediately,
starting from the next iteration of the loop, as the machine code represents
exactly the loop that was being interpreted so far.

This process assumes that the path through the loop that was traced is a
``typical'' example of possible paths. Of course
it is possible that later another path through the loop is taken, in which case
one of the guards that were put into the machine code will fail.\footnote{There are more
complex mechanisms in place to still produce more code for the cases of guard
failures \cite{andreas_gal_incremental_2006}, but they are independent of the issues discussed in this
paper.}

>>>>

@@

When a hot loop is identified, the interpreter enters a special mode, called
\emph{tracing mode}. During tracing, the interpreter records a history of all
the operations it executes. It traces until it has recorded the execution of one
iteration of the hot loop. To decide when this is the case, the trace is
repeatedly checked during tracing as to whether the interpreter is at a position
in the program where it had been earlier.

The history recorded by the tracer is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce efficient machine code by generating
code from the operations in it.  The machine code can then be executed immediately,
starting from the next iteration of the loop, as the machine code represents
exactly the loop that was being interpreted so far.

Being sequential, the trace represents only one
of the many possible paths through the code. To ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example conditions and indirect or virtual
calls.  When generating the machine code, every guard is turned into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit the machine code and continue the execution by falling
back to interpretation.\footnote{There are more complex mechanisms in place to
still produce more code for the cases of guard failures
\cite{andreas_gal_incremental_2006}, but they are independent of the issues
discussed in this paper.}
